Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-12-09 analysis/paper/paper.Rmd
some TB edits to Methods

This research is an essential first step in characterizing conditions through the Deadwater Reach and providing data necessary to discuss potential predation mitigation strategies (if needed). We follow with a discussion of the various assumptions that went into the mark-recapture and bioenergetics models and how violations of some assumptions may affect overall results and inferences from the study. We also consider how consumption of juvenile Chinook Salmon by predators at Deadwater Slough might impact adult returns, hindering recovery of populations in the Upper Salmon River MPG.

<!-- where Northern Pikeminnow removal has proven fruitful in increasing survival through the reservoir systems [@Winther2019]. -->

@@

Our first objective was to estimate the population size of piscine predators in the Deadwater Slough to begin to understand their potential impacts on juvenile Chinook Salmon. The intent was to estimate predator abundances using mark-recapture methods during fall 2019 and spring 2020 during emigration of the 2018 brood year. We attempted to sample near the peaks of the juvenile DSR (fall) and NRR (spring) emigrations. However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 sampling event was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional sampling event was added for spring 2021. The study design was such that the fall 2019 and fall 2020 sampling efforts included both a "mark week" and "recapture week" intended for mark-recapture models. However, due to budget constraints the spring 2021 effort only included a single "mark week"; instead, we calculated the catch-per-unit-effort (CPUE; fish per hour) for the spring 2021 sampling event for comparison to the fall efforts. In the end, sampling dates were as follows: fall 2019 (mark = November 12-14; recapture = November 19-21), fall 2020 (mark = October 20-23; recapture = October 27-30); spring 2021 (mark = May 18-21). The recapture events were held just one week following the mark events to reduce opportunity for immigration/emigration of predators between events i.e., to avoid violation of a closed population for mark-recapture models.

We attempted a multiple of capture methods during the first year including raft electrofishing, fyke netting, snorkeling, and angling to reduce selectivity, both for species and size classes. Unfortunately, some methods proved ineffective during initial efforts. For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. And due to debris floating in the water column, there were concerns with fyke nets being pulled from their anchored locations and becoming entangled. Additionally, there were concerns with raft electrofishers and fyke nets incidentally capturing ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. Eventually, angling proved to be the most effective method for capturing piscine predators while minimizing potential impacts to steelhead but recognizing that angling can have both species and size selectivity (bias). From here on, we focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were allowed to fish anywhere within Deadwater Slough; periodically or upon filling a livewell anglers would boat or hike their catch to a processing station located at the boat ramp approximately 500 m downstream from the top of the slough. For each fish, processors would record the date of capture, species, total length (TL; mm), and whether the fish was previously marked. If not previously marked, each fish was provided a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique for each marking day. We also recorded whether marked individuals were released back to be captured again. For release, processors attempted to distribute released individuals throughout Deadwater Slough to facilitate mixing back into the population. We also recorded the crew (person or combination of persons) that captured each fish and the start and end times for each crew to estimate total effort.

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon, other fishes (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner *Richardsonius balteatus*, etc.), and non-targets (e.g., macroinvertebrates, organic matter, etc.). Stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks immediately following lavage for later analysis in a controlled environment. For each sample, wet weight (grams) was recorded for the total combined stomach content as well as the contents consisting of fish parts/matter. Fish and fish remnants were identified to the lowest taxonomic unit, when possible, or were categorized as unknown. A subset of Northern Pikeminnow captures were euthanized for dissection after gastric lavage to validate the efficacy of the methodology.

@@

Gastric lavage was used in an attempt to document predation of juvenile Chinook Salmon by Northern Pikeminnow. However, we were ultimately interested in the potential total number or amount of juveniles consumed during the peaks of DSR (fall) and NRR (spring) emigrations. To estimate the consumption of juvenile Chinook Salmon in the Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 and applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon based on an estimate from @Moss2016. We used a weight-length formula [@Parker1995] to calculate the average predator start weight, converted from the average length of Northern Pikeminnow captured during our study and using the FSA package [@Ogle2021] in R. Continuous water temperature data from 2019 was used in two alternative models, one using the baseline start and end weights and the second with a 10% increase in average end weight on Northern Pikeminnow.

We ran three separate bioenergetic models for this study. For the first model, we used temperature data from a 78-day period of September 1 through November 17 where DSR emigrants are known to enter the mainstem Salmon River from natal tributaries (e.g., Lemhi River) and begin their downstream migration.<!-- Did we just use data from Lemhi here? All tributaries? Was this subjective? --> During this time, water temperatures typically exceed the range (0-7$^\circ$C) that would illicit concealment behavior or torpor from juvenile Chinook Salmon<!--citation?-->. This first model assumed no growth for Northern Pikeminnow. The second model was instead run for one full year, but also assumed no growth for Northern Pikeminnow. The assumption of zero growth was included to estimate consumption if either individual Northern Pikeminnow were not growing and/or the population is stable. Finally, the alternative third model shows how an increase of 10% body weight or population size changes the estimates of consumption over a full year period.

The model runs provided us with an estimate of the total biomass consumed by an individual Northern Pikeminnow for the given time period. We then multipled biomass consumed by an individual Northern Pikeminnow by the estimated population size for that time period to get an estimate of the total biomass consumed. The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed by dividing by a eight of 12 g per individual. This weight was an average from juvenile Chinook Salmon emigrating past six rotary screw traps operating in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.

>>>>

This research is an essential first step in characterizing conditions through the Deadwater Reach and providing data necessary to discuss potential predation mitigation strategies (if needed), where Northern Pikeminnow removal has proven fruitful in increasing survival through the reservoir systems [@Winther2019]. We follow with a discussion of the various assumptions that went into the mark-recapture and bioenergetics models and how violations of some assumptions may affect overall results and inferences from the study. We also consider how consumption of juvenile Chinook Salmon by predators at Deadwater Slough might impact adult returns, hindering recovery of populations in the Upper Salmon River MPG.

@@

Our first objective was to estimate the population size of piscine predators in the Deadwater Slough using mark-recapture methods. We attempted to sample predators near the peaks of the juvenile DSR and NRR emigrations for the 2018 brood year (fall 2019 and spring 2020, respectively). However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 effort was moved to fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional effort was added in spring 2021. Each of the fall efforts included sampling across a two-week span; however, due to budget constraints the spring 2021 effort only included a single week. For the fall efforts, the first week was considered a "mark event" and the second week a "recapture event" for single-census mark-recapture estimators (described below); therefore, the "recapture week" occurred the week following the "mark week" to minimize predator emigration/immigration and constrain the population for the mark-recapture models.

<!-- In the end, sampling dates were as follows: fall 2019 (mark = November 12-14; recapture = November 19-21), fall 2020 (mark = October 20-23; recapture = October 27-30); spring 2021 (mark = May 18-21). -->

Multiple capture methods were employed during the fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. Unfortunately, some methods proved ineffective during initial efforts. For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. Eventually, angling proved to be the most effective method for capturing piscine predators while minimizing potential impacts to steelhead but recognizing that angling can have both species and size selectivity (bias). From here on, we focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were permitted to fish anywhere within Deadwater Slough. During a sample event, anglers would boat or hike their catch (periodically or upon filling a livewell) to a processing station located at the boat ramp approximately 500 m downstream from the top of the slough. For each fish, we recorded the date of capture, species, total length (TL; mm), and whether the fish was previously marked. Unmarked fish were given a unique physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) each day and released. Released individuals were distributed throughout Deadwater Slough to facilitate mixing back into the population. Finally, we recorded the angling start and end times for each crew (person or combinations of persons) to facilitate calculations of catch per unit effort (CPUE).

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner *Richardsonius balteatus*, etc.). Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks for later analysis in a controlled environment. For each sample, wet weight (grams) was recorded for the total combined stomach content, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as the portion consisting of fish parts/matter. Fish and fish remnants were identified to the lowest taxonomic unit, when possible, or were categorized as unknonw. A subset of Northern Pikeminnow captures (~ 5%) were euthanized for dissection after gastric lavage to validate the efficacy of the methodology.

@@

To estimate the total consumption potential (i.e., number/amount of juvenile consumed during the peaks of DSR [fall] and NRR [spring] emigrations) of Northern Pikeminnow in Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon based on an estimate from @Moss2016. We used a weight-length formula [@Parker1995] to calculate the average predator start weight from the average length of Northern Pikeminnow captured during our study using the FSA package [@Ogle2021] in R. Continuous water temperature data from 2019 was used in two alternative models, one using the baseline start and end weights and the second with a 10% increase in average end weight on Northern Pikeminnow.

We ran three separate bioenergetic models for this study. For the first model, we used temperature data from a 78-day period spanning September 1 through November 17 when DSR emigrants are known to enter the mainstem Salmon River from natal tributaries (e.g., Lemhi River) and begin their downstream migration.<!-- Did we just use data from Lemhi here? All tributaries? Was this subjective? --> During this time, water temperatures typically exceed the range (0-7$^\circ$C) that would illicit concealment behavior or torpor from juvenile Chinook Salmon<!--citation?-->. This first model assumed no growth for Northern Pikeminnow. The second model was instead run for one full year, but also assumed no growth for Northern Pikeminnow. The assumption of zero growth was included to estimate consumption if individual Northern Pikeminnow were not growing and/or the population is stable. Finally, the third model shows how an increase of 10% body weight or an increase in population size changes the estimates of consumption over a full year period.

To estimate the total biomass consumed, we then multipled biomass consumed by an individual Northern Pikeminnow for the given time period by the estimated population size for that time period. The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed by dividing by a weight of 12 g per individual. This weight was an average from juvenile Chinook Salmon emigrating past six rotary screw traps operating in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.
